Tulai

### 2023-09-15 paper/paper_tulai.qmd
some editions on WAIB and PCA plots.

@@

@@

```{r}
#Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
#I thought this format would simplify the process of those depths needed to be splitting up.

@@

sum_depth_sheet <- read_sheet(google_sheet_url)

@@

# Separate `excavation area` into `length` and `width` and calculate `volume`
sum_depth_sheet <- sum_depth_sheet %>%
  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
  mutate(
    length = as.numeric(length),
    width = as.numeric(width),
    `thickness of deposit` = `thickness of deposit` / 100  # Convert from cm to m
  ) %>%
  mutate(
    volume = length * width * `thickness of deposit`  # Calculate volume in m^3
  )

@@

```{r}
#WABI plot: here I do not consider those levels with a value 0 of retouch frequency.

library(ggpmisc)

# Filter the data (get out those level with Retouch frequency=0)
filtered_data <- sum_depth_sheet %>% filter(lithic_dens > 0 & retouch_freq > 0)

# Create the plot: WABI_Level_logscale
WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_dens,
                y = retouch_freq)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels 
  geom_text(aes(label = paste(area, layer, level, sep = "_")), 
            vjust = 1.5, 
            hjust = 0.5, 
            check_overlap = TRUE) +

@@

#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, should we not consider it?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index); 
##core frequency
##blank frequency
##retouch frequency
##tool diversity

@@

# calculate Tool Diversity
##calculating tool diversity is a bit tricky here; as I have some spitted depth, which we assign the number of artefacts based on the calculation of splitting, so how should I consider the tool diversity here?
```

>>>>

    ##SS: this column was so messy with too many mistakes; I needed to have all typos correct then transfer them in their corresponding columns.  

@@

library(dplyr)
library(stringr)

# Define the patterns to search for
patterns <- c("denticulate-notch", "retouched piece", "scraper",
              "Perforator", "serrated scraper", "backed pieces",
              "truncated pieces", "burin", "geometric", "scraper-notch",
              "microburin")

# Update the `retouch` column based on the `tooltypegroup` column
tl_final <- tl_final %>%
  mutate(retouch = ifelse(str_detect(tooltypegroup, str_c(patterns, collapse = "|")), 1, retouch))

# View the first few rows to confirm
head(tl_final)
rev(sort(table(tl_final$tooltypegroup)))
rev(sort(table(tl_final$retouch)))
```

@@

```{r}
# Generate a summary table by grouping by 'depth'
summary_table_depth_tool_type <- tl_final %>%
  group_by(depth) %>%
  summarise(tool_type = n_distinct(tooltypegroup))

# Create a full summary table by joining it back to your original summary table based on 'depth'
summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
  left_join(summary_table_depth_tool_type, by = "depth")

# Print column names 
print(names(summary_table_depth_area_layer))

kable(summary_table_depth_area_layer, 
      caption = "Table: Summary by Depth, Area, and Layer", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

```{r}
# Specify the exact columns want to keep
desired_columns <- c("TA", "depth", "area", "layer", "upper", "lower", 
                     "Cores (n)", "Core rejuvenations pieces (n)", 
                     "Blade (n)", "Flake (n)", "Bladelet (n)", 
                     "Retouched Tools (n)", "Utilized (n)", "Sickle Shine (n)",
                     "tool_type")  # Only keep this 'tool_type'

# Filter the tl_final dataframe
summary_table_depth_area_layer <- summary_table_depth_area_layer %>%
  filter(area %in% c('ta', 'd1', 'tp1'))

kable(summary_table_depth_area_layer, 
      caption = "Table: Summary by Depth, Area, and Layer", 
      align = 'c') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12) %>%
  row_spec(0, bold = TRUE, font_size = 14)

#Here, I've imported the table 'Summary by Depth, Area, Layer' into Google Sheets for easier management.
#I thought this format would simplify the process of those depths needed to be splitting up.

@@

# Create the 'label' column
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(label = paste(area, layer, level, sep = "_"))

# Add a suffix only to duplicates
sum_depth_sheet <- sum_depth_sheet %>%
  group_by(label) %>%
  mutate(suffix = case_when(
    n() == 1 ~ NA_character_,
    TRUE ~ letters[row_number()]
  )) %>%
  ungroup()

# Append the suffix to 'label'
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(label = if_else(is.na(suffix), label, paste0(label, "_", suffix))) %>%
  select(-suffix)

# Relocate 'label' column after 'level'
sum_depth_sheet <- sum_depth_sheet %>%
  relocate(label, .after = level)

@@

#### previously I had a column'excavation area'; I seprated them in 2 columns in order to calculate the colume; but now when I wanted to run the code it said "! object 'excavation area' not found"; would it be OK to delete these lines?

# Separate `excavation area` into `length` and `width` and calculate `volume`
sum_depth_sheet <- sum_depth_sheet %>%
  separate(`excavation area`, into = c("length", "width"), sep = "\\*") %>%
  mutate(
    length = as.numeric(length),
    width = as.numeric(width),
    `thickness` = `thickness` / 100  # Convert from cm to m
  ) %>%
  mutate(
    volume = length * width * `thickness`  # Calculate volume in m^3
  )

@@

```{r}
# Create the plot: WABI_Level_logscale
# here I do not consider those levels with a value 0 of retouch frequency. 
WABI_Level <- ggplot(filtered_data,  
            aes(x = lithic_dens,
                y = retouch_freq)) +
  geom_point(color = "black", size = 1.5) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              linewidth = 0.5) +
  # Add text labels based on 'label' column
  geom_text(aes(label = label), 
            vjust = 1.5, 
            hjust = 0.5, 
            check_overlap = TRUE) +

@@

#PCA_level: for PCA Bicho and Cascalheira considered these variables: estimated area (as we calculated based on levels, I consider the thickness; is that OK?), core frequency, blank frequency, chip frequency: artefacts smaller than 1 cm (as we work on a neolithic asssemblage, we have so many small artefacts; so, I do not consider it), feature frequency (we have some stone alignments, but we do not know the exact level, we know the area of those features), retouch frequency, tool diversity (diversity of tool types within each assemblage, calculated using Menhinick’s index); 
##core frequency
##blank frequency
##retouch frequency
##tool diversity
##thickness of deposit
##lithic density

@@

# Calculate tool diversity using Menhinick's index
sum_depth_sheet <- sum_depth_sheet %>%
  mutate(
    total_tools = retouch_tool + utilized + sickle_shine,
    tool_diversity = ifelse(total_tools > 0, n_distinct(tool_type) / sqrt(total_tools), NA)
  )

```{r}
pca_data <- sum_depth_sheet %>% 
  select(thickness, lithic_dens, core_freq, blank_freq, retouch_freq, tool_diversity) %>%
  na.omit()  # Remove rows with NA values

```{r}
library(FactoMineR)
library(factoextra)
# Assuming 'sum_depth_sheet' is your dataframe and you have added 'm_index' as your Menhinick index
pca_data <- sum_depth_sheet[, c('thickness', 'lithic_dens', 'core_freq', 'blank_freq', 'retouch_freq', 'tool_diversity')]

# Perform PCA
res.pca <- PCA(pca_data, graph = FALSE)

# View the summary results
summary(res.pca)

library(kableExtra)
library(htmltools)

# Create a new column named "Dimension" 
eigen_df$Dimension <- paste("Dim", seq_len(nrow(eigen_df)), sep = "")

eigen_df <- eigen_df[, c("Dimension", "eigenvalue", "percentage of variance", "cumulative percentage of variance")]

# Round the numerical columns to 3 decimal places
eigen_df[, 2:4] <- round(eigen_df[, 2:4], 3)

# Creating the kable output
kable_output <- kable(eigen_df, 
                      "html", 
                      align = 'c', 
                      col.names = c("Dimension", 
                                    "Eigenvalue", 
                                    "Variance Percent", 
                                    "Cumulative Variance Percent")) %>%
  kable_styling("striped", 
                full_width = F) %>%
  add_header_above(c(" " = 1, "Eigenvalues and percentage of variance for each dimension of PCA" = 3))

# Display the table
htmltools::browsable(kable_output)

```{r}
# Visualize biplot:1
fviz_pca_biplot(res.pca,
                axes = c(1, 2))
rownames(res.pca$ind$coord) <- sum_depth_sheet$label
# Visualize biplot with row names
fviz_pca_biplot(res.pca)

```{r}
# Visualize biplot: 2
fviz_pca_biplot(res.pca,
                axes = c(3, 4))
rownames(res.pca$ind$coord) <- sum_depth_sheet$label
# Visualize biplot with row names
fviz_pca_biplot(res.pca)

-----------------------------------------------
DONE UPTO HERE
-----------------------------------------------
